Model Checking Differentially Private Properties

### 2018-01-26 semantics.tex
try

@@

\paragraph{Advanced Compositional Properties}

One of the advantage of applying model checking is that we may get tighter bounds for compositional properties. Consider ...
\lz{find a -- simpler -- formula such that the bound is not tight, maybe $\delta>0$ is needed}

@@

\section{Markov Decision Processes}
\lz{let put it as another section after Markov chains}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   position-independent scheduler } \\
  & & \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
  \\
  & &
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}
Recall that $M_{\scheduler{S}}$ is but a Markov chain. The semantics
of $M_{\scheduler{S}}, \neighbor{S}, \pi \models \phi$ and hence the
probability $\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}$ are
defined as in Markov chains.
The semantics of $\dpCTL$ on reactive Markov decision processes
is again standard except the differentially private operator
$\dpriv{\epsilon}{\delta}$. For any path formula $\phi$,
$\dpriv{\epsilon}{\delta} \phi$ specifies states which are $(\epsilon,
\delta)$-close to all its neighbors in having paths satisfying $\phi$
for arbitrary position-independent schedulers. That is, no
position-independent scheduler can force any of neighbors to have
probabilistically distinguishable behaviors.

>>>>

\section{Model Checking Algorithms}
The model checking algorithms for $\dpCTL$ follow the classical
algorithms for $\PCTL$ by computing the states satisfying
sub state-formulae inductively~\cite{BK:08:PMC}. It hence suffices to
consider the inductive step where the states satisfying the subformula
$\dpriv{\epsilon}{\delta} (\phi)$ is computed.

%We start with the algorithm for Markov chains.

In the classical $\PCTL$ model checking algorithm for Markov chains,
states satisfying the subformula $\PJ{J} \phi$ are obtained by
computing $\myPr{s}{K}{\neighbor{S}}{\phi}$ for state $s \in S$.
These probabilities can be computed by solving linear equations or
through iterative approximations. We summarize it in the following
theorem~\cite{BK:08:PMC}:

\begin{lemma}
  Let $K = (S, \wp, L)$ be a Markov chain, $s
  \in S$, and $B, C \subseteq S$. The probabilities
  $\myPr{s}{K}{\neighbor{S}}{\gX B}$ and
  $\myPr{s}{K}{\neighbor{S}}{B \mho C}$
\hide{
  $\Pr[\{ \pi : K, \neighbor{K}, \pi \models B \buntil{n} C \textmd{
    with } \pi_0 = s \}]$
} are computable within time polynomial in
  $|S|$.
  \label{lemma:PJ-subroutines}
\end{lemma}

In Lemma~\ref{lemma:PJ-subroutines}, we abuse the notation slightly to
admit path formulae of the form $\gX B$ (next $B$) and
$B \guntil C$ ($B$ until $C$) with $B, C \subseteq S$ as
in~\cite{BK:08:PMC}. They are interpreted by introducing new atomic
propositions $B$ and $C$ for each $s \in B$ and $s \in C$
respectively. We will also use graphical symbols $\gF C$ ($C$ in
the future) and $\gG C$ (globally $C$) in such path formulae
$\varphi$ to minimize confusion.

In order to determine the set $\{ s : K, \neighbor{S}, s \models
\dpriv{\epsilon}{\delta} \varphi \}$, our algorithm first computes
the probabilities $p (s) = \myPr{s}{K}{\neighbor{S}}{\varphi}$ for every
$s \in S$ (Algorithm~\ref{algorithm:sat-dpriv-dtmc}). For each $s \in S$,
it then compares the probabilities $p (s)$ and $p (t)$ for every
neighbor $t$ of $s$. If there is a neighbor $t$ such that $p (s)$ and
$p (t)$ are not $(\epsilon, \delta)$-close, the state $s$
is removed from the result. Hence
Algorithm~\ref{algorithm:sat-dpriv-dtmc} returns all states which are
$(\epsilon, \delta)$-close to their neighbors.
The algorithm requires at most $O
(|S|^2)$ additional steps. We hence have the following results:

\begin{algorithm}
  \begin{algorithmic}[1]
    \Function{SAT}{$K$, $\neighbor{S}$, $\phi$}
    \Match{$\phi$}
    \Comment{by Lemma~\ref{lemma:PJ-subroutines}}
    \Case{$\X \Psi$}
      \State{$B \leftarrow \textmd{SAT} (K, \neighbor{S}, \Psi)$}
      \State{$p(s) \leftarrow \myPr{s}{K}{\neighbor{S}}{\gX B}$
        for every $s \in S$}
    \EndCase
    \Case{$\Psi \until \Psi'$}
      \State{$B \leftarrow \textmd{SAT} (K, \neighbor{S}, \Psi)$}
      \State{$C \leftarrow \textmd{SAT} (K, \neighbor{S}, \Psi')$}
      \State{$p(s) \leftarrow \myPr{s}{K}{\neighbor{S}}{B \guntil C}$ for
        every $s \in S$}
    \EndCase
    \EndMatch
    \State{$R \leftarrow S$}
    \For{$s \in S$}
      \For{$t$ with $s \neighbor{K} t$}
        \If{$p(s) \not\leq \epsilon p(t) + \delta$ or
            $p(t) \not\leq \epsilon p(s) + \delta$}
        {remove $s$ from $R$}
        \EndIf
      \EndFor
    \EndFor

    \State{\Return $R$}
    \EndFunction
  \end{algorithmic}
  \caption{SAT($\phi$, $\neighbor{K}$)}
  \label{algorithm:sat-dpriv-dtmc}
\end{algorithm}

\begin{proposition}
  Let $K = (S, \wp, L)$ be a Markov chain.
  $\{ s : K, \neighbor{K}, s \models \dpriv{\epsilon}{\delta} \phi \}$ is
  computable within time polynomial in $|S|$.
\end{proposition}

\begin{corollary}
  Let $K = (S, \wp, L)$ be a Markov chain and $\Phi$ a $\dpCTL$
  formula. $\{ s : K, \neighbor{K}, s \models \Phi \}$ is
  computable within time polynomial in $|S|$ and $|\Phi|$.
\end{corollary}

@@

\paragraph{Advanced Compositional Properties}
One of the advantage of applying model checking is that we may get tighter bounds for compositional properties. Consider our survey example, and the property $\dpriv{1}{.5} (\X Y)$. Obviously, it holds in states $+$ and $-$. A careful check ensures that one cannot decrease the value of $\epsilon_1=1$ or $\delta_1=.5$ without increasing the other one.

Now we consider the formula $\dpriv{\epsilon_2}{\delta_2} (\X (Y\wedge \X Y))$ under the model in ???. Applying the compositional theorem\lz{Bow-Yaw, as we only refer to the book, I wonder whether using the notion $e^\epsilon$ is easier to explain?}, one can choose $\epsilon_2=\epsilon_1^2=1$ and $\delta_2=2\delta_1=1$. However, we can check easily that one can get better privacy parameter $(1,0.5)$ or $(0,\frac{9}{16})$.

@@

\section{Markov Decision Processes}
The semantics over Markov chains generalizes to Markov decision
processes as well. Note that two states with different enabled actions
are trivially distinguishable; no privacy can be preserved. Without
loss of generality, we consider reactive Markov decision processes.
Let $M = (S, \Act, \wp, L)$ be a reactive Markov
decision process. Define the satisfaction
relation $M, \neighbor{S}, s \models \Phi$ as follows.
\begin{eqnarray*}
  M, \neighbor{S}, s \models \top\\
  M, \neighbor{S}, s \models p
  & \textmd{ if } &
  p \in L(s)\\
  M, \neighbor{S}, s \models \neg \Phi
  & \textmd{ if } &
  M, \neighbor{S}, s \not\models \Phi\\
  M, \neighbor{S}, s \models \Phi_0 \wedge \Phi_1
  & \textmd{ if } &
  M, \neighbor{S}, s \models \Phi_0 \textmd{ and }
  M, \neighbor{S}, s \models \Phi_1\\
  M, \neighbor{S}, s \models \PJ{J} \phi
  & \textmd{ if } &
  \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \in J
  \textmd{ for every scheduler } \scheduler{S}\\
  M, \neighbor{S}, s \models \dpriv{\epsilon}{\delta} \phi
  & \textmd{ if } &
  \textmd{for all } t \textmd{ with } s \neighbor{S} t \textmd{ and
   position-independent scheduler } \\
  & & \scheduler{S},
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
   \textmd{ and }
   \myPr{t}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} \leq \epsilon \cdot
  \\
  & &
   \myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi} + \delta
\end{eqnarray*}

\lz{how about using "query sequence" instead of position-independent scheduler, and then argue that it makes sense for the resulting MDP model from differential privacy? This corresponds to a special scheduler that is applied on both neighboring states. }
Recall that $M_{\scheduler{S}}$ is but a Markov chain. The semantics
of $M_{\scheduler{S}}, \neighbor{S}, \pi \models \phi$ and hence the
probability $\myPr{s}{M_{\scheduler{S}}}{\neighbor{S}}{\phi}$ are
defined as in Markov chains.
The semantics of $\dpCTL$ on reactive Markov decision processes
is again standard except the differentially private operator
$\dpriv{\epsilon}{\delta}$. For any path formula $\phi$,
$\dpriv{\epsilon}{\delta} \phi$ specifies states which are $(\epsilon,
\delta)$-close to all its neighbors in having paths satisfying $\phi$
for arbitrary position-independent schedulers. That is, no
position-independent scheduler can force any of neighbors to have
probabilistically distinguishable behaviors.

### 2018-01-26 syntax.tex
first try
added some properties
try

Consider
\begin{eqnarray*}
  \Phi & ::= & \top \ |\ p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\ 
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \X \Phi \ |\ \Phi \until \Phi
\end{eqnarray*}
A \emph{state} formula $\Phi$ is either $\top$, an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A 
\emph{path} formula $\phi$ is a \emph{next} operator ($\X$)
followed by a state formula, or an \emph{until} operator
($\until$) operator enclosed by two state formulae.

>>>>

Consider
\begin{eqnarray*}
  \Phi & ::= & \top \ |\ p \ |\ \neg \Phi \ |\ \Phi \wedge \Phi \ |\
               \PJ{J} \phi \ |\ \dpriv{\epsilon}{\delta} \phi\\
  \phi & ::= & \Phi \ |\ \neg\phi  \ |\  \phi\wedge\phi  \ |\  \gX \phi \ |\ \phi \until \phi
\end{eqnarray*}
\lz{Bow-Yaw: we use $\X$ or $\gX$?}
A \emph{state} formula $\Phi$ is either $\top$, an atomic proposition
$p$, the negation of a sate formula, the conjunction of two state
formulae, a \emph{probabilistic} operator $\PJ{J}$ with $J$
an interval in $[0, 1]$ followed by a path formula, or a
\emph{differentially private} operator $\dpriv{\epsilon}{\delta}$
 with two non-negative real numbers $\epsilon$ and $\delta$ followed
 by a path formula. A
\emph{path} formula $\phi$ is a \emph{next} operator ($\X$)
followed by a state formula, or an \emph{until} operator
($\until$) operator enclosed by two state formulae.
